---
title: "RNGfroGPD Vignette"
author: "Hesen Li, Ruizhe Chen, Hai Nguyen, Yu-che Chung, Ran Gao, Hakan Demirtas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RNGfroGPD_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE, include=FALSE}
library(RNGforGPD)
library(corpcor)
library(VGAM)
library(mvtnorm)
library(Matrix)
```

# Introduction

This vignette file conveys certain ideas behind the generalized Poisson distribution and some examples of applying the functions in this package (**RNGforGPD**).


# Functions and Comments


**GenUniGpois**

We choose different data generation methods according to different parameter values because restrictions apply when the rate parameter or the dispersion parameter of the generalized Poisson is within certain ranges. For example, the normal approximation method does not work well for theta < 10. 


```{r, echo = FALSE}
GenUniGpois = function(theta, lambda, n, details = TRUE, method){
  # Check if parameters are valid
  tol = .Machine$double.eps^0.5
  if (n <= 0 | abs(n - round(n)) > tol) {
    stop("n must be a positive integer.")
  }
  if(n == 1) stop("Parameter estimates are exactly the same as input of parameters in the case when n = 1!")
  if(theta <= 0) stop("Theta has to be greater than 0!")
  if(lambda >= 1) stop("Lambda has to be less than 1!")
  if (lambda < 0 & lambda < (-theta) / 4) stop(paste("For lambda < 0, lambda must be >= -theta/4 = ", (-theta)/4, "!", sep = ""))
  myset = numeric(n)
  if (lambda == 0) { # 1. rpois function
    myset = rpois(n,theta)
    if (n != 1) {
      model = vglm(myset ~ 1, family = genpoisson(zero = 1))
      coef = Coef(model, matrix = TRUE)
      emp.theta = round(coef['theta'], 6)
      emp.lambda = round(coef['lambda'], 6)
    }
  } else if (method == "Inversion"){ # 2. Inversion Method (when lambda is negative)
    w = exp(-lambda)
    for (i in 1:n){
      mys = exp(-theta)
      myp = mys
      x = 0
      u = runif(1)
      while (u > mys){
        x = x + 1
        myc = theta - lambda + lambda*x
        myp = w * myc * (1 + lambda/myc)^(x-1) * myp * x^(-1)
        mys = mys + myp
      }
      myset[i] = x
    }
    if (n != 1) {
      model = vglm(myset ~ 1, family = genpoisson(zero = 1))
      coef = Coef(model, matrix = TRUE)
      emp.theta = round(coef['theta'], 6)
      emp.lambda = round(coef['lambda'], 6)
    }
  } else if (method == "Branching") { # 3. Branching Method (when lambda is positive)
    if (lambda < 0) stop("Lambda should be greater than 0!")
    for (i in 1:n){
      index = 0
      y = rpois(1,theta)
      x = y
      if (y <= 0) myset[i] = x
      else while(index == 0){
        z = rpois(1, lambda * y)
        x = x + z
        y = z
        index = 1 * (y <= 0)
      }
      myset[i] = x
    }
    if (n != 1) {
      model = vglm(myset ~ 1, family = genpoisson(zero = 1))
      coef = Coef(model, matrix = TRUE)
      emp.theta = round(coef['theta'], 6)
      emp.lambda = round(coef['lambda'], 6)
    }
  } else if (method == "Normal-Approximation") { # 4. Normal-Approximation Method (when theta is large enough which is bigger than 10)
    if (theta < 10) warning("Normal approximation may not be reliable for theta less than 10!")
    mym = theta / (1 - lambda)
    myv = sqrt(theta * (1 - lambda)^(-3))
    y = rnorm(n)
    x = floor(mym + myv * y + 0.5)
    x[x < 0] = 0
    myset = x
    if (n != 1) {
      model = vglm(myset ~ 1, family = genpoisson(zero = 1))
      coef = Coef(model, matrix = TRUE)
      emp.theta = round(coef['theta'], 6)
      emp.lambda = round(coef['lambda'], 6)
    }
  } else if (method == "Build-Up") { # 5. Build-Up Method
    mynumx = numeric(n)
    t = exp(-theta)
    for (i in 1:n) {
      u = runif(1)
      x = 0
      px = t
      s = px
      while (u > s) {
        x = x + 1
        px = theta * (theta + lambda * x)^(x - 1)*exp(-theta - lambda * x) / factorial(x)
        s = s + px
      }
      myset[i] = x
    }
    if (n != 1) {
      model = vglm(myset ~ 1, family = genpoisson(zero = 1))
      coef = Coef(model, matrix = TRUE)
      emp.theta = round(coef['theta'], 6)
      emp.lambda = round(coef['lambda'], 6)
    }
  } else if (method == "Chop-Down") { # 6. Chop-Down Method
    mynump = numeric(n)
    t = exp(-theta)
    for (i in 1:n) {
      u = runif(1)
      x = 0
      px = t
      while (u > px) {
        u = u - px
        x = x + 1
        px = theta * (theta + lambda * x)^(x - 1) * exp(-theta - lambda * x) / factorial(x)
      }
      myset[i] = x
    }
    if (n != 1) {
      model = vglm(myset ~ 1, family = genpoisson(zero = 1))
      coef = Coef(model, matrix = TRUE)
      emp.theta = round(coef['theta'], 6)
      emp.lambda = round(coef['lambda'], 6)
    }
  }
  if (details == TRUE) {
    print(paste("Specified theta is ", theta, ", empirical theta is ",
                 emp.theta, ", specified lambda is ", lambda, ", empirical lambda is ", emp.lambda, ".", sep = ""))
  }
   return(invisible(list(data = myset,
                         specified.theta = theta,
                         empirical.theta = as.numeric(emp.theta),
                         specified.lambda = lambda,
                         empirical.lambda = as.numeric(emp.lambda))))
}
```


```{r}
GenUniGpois(2, 0.9, 100, method = "Branching")
GenUniGpois(5, -0.4, 100, method = "Inversion")
GenUniGpois(12, 0.5, 100, method = "Normal-Approximation")
data = GenUniGpois(10, 0.4, 10, method = "Chop-Down", details = FALSE)
data = GenUniGpois(3, 0.9, 10000, method = "Build-Up", details = FALSE)
```


**ComputeCorrGpois**

From a practical perspective, correlation bounds among variables are typically narrower than between −1 and 1 (the theoretical maximum and minimum correlation bounds) because different correlation upper and lower bounds may be imposed by the marginal distributions. A simple sorting technique can be used to obtain approximate correlation bounds and this approach works regardless of the data type or the distributional assumption (Demirtas, Hedeker 2011).

Using the sorting technique, we wrote the function `ValidCorrGpois` that computes the lower and upper correlation bounds between a pair of generalized Poisson variables. Besides, this function serves as an integral part of the `ValidCorrGpois` function, which examines whether values of pairwise correlation matrices fall within the limits imposed by the marginal distributions.


```{r, echo = FALSE, warning=FALSE}
ComputeCorrGpois = function(theta.vec, lambda.vec) {
  no.gpois = length(theta.vec)
  samples = 1e+05
  u = matrix(NA, nrow = no.gpois, ncol = samples)
  for (i in 1:no.gpois){
    u[i,] = GenUniGpois(theta.vec[i], lambda.vec[i], samples, method = "Inversion", details = FALSE)$data
  }
  maxmat = minmat = diag(NA, no.gpois)
  errorCount = 0
  for (i in 1:no.gpois){
    for (j in 1:no.gpois){
      if (i != j) {
        maxcor = cor(sort(u[i,]), sort(u[j,]))
        mincor = cor(sort(u[i,]), sort(u[j,], decreasing = TRUE))
        minmat[i, j] = mincor
        maxmat[i, j] = maxcor
        cat(".")
      }
    }
  }
  cat("\n")
  return(list(min = minmat, max = maxmat))
}
```


```{r}
ComputeCorrGpois(c(3,2,5,4),c(0.3,0.2,0.5,0.6))
ComputeCorrGpois(c(4,5),c(-0.45,-0.11))
```


**ValidCorrGpois**

This function checks the required conditions of the values of pairwise correlations, which include positive definiteness, symmetry, correctness of dimensions, and whether the correlations fall within the correlation bounds. `ValidCorrGpois` ensures that the supplied correlation matrix is valid for simulating multivariate generalized Poisson distributions using `GenMVGpois`. 


```{r, echo = FALSE}
ValidCorrGpois = function (corMat, theta.vec, lambda.vec) {
  no.gpois = length(theta.vec)
  errorCount = 0
  if (ncol(corMat) != (no.gpois)) {
    stop("Dimension of correlation matrix does not match the number of variables!\n")
  }
  if (is.positive.definite(corMat) == FALSE) {
    stop("Specified correlation matrix is not positive definite! \n")
  }
  if (isSymmetric(corMat) == FALSE) {
    stop("Specified correlation matrix is not symmetric! \n")
  }
  if (sum(corMat > 1) > 0) {
    stop("Correlation values cannot be greater than 1! \n")
  }
  if (sum(corMat < (-1)) > 0) {
    stop("Correlation values cannot be less than -1! \n")
  }
  if (sum(diag(corMat) != 1) > 0) {
    stop("All diagonal elements of the correlation matrix must be 1! \n")
  }
  maxcor = ComputeCorrGpois(theta.vec, lambda.vec)$max
  mincor = ComputeCorrGpois(theta.vec, lambda.vec)$min
  for (i in 1:nrow(corMat)) {
    for (j in 1:i) {
      if (errorCount == 0)
        cat(".")
      if (i != j) {
        if (corMat[i, j] > maxcor[i, j] | corMat[i, j] <  mincor[i, j]) {
          cat("\n corMat[", i, ",", j, "] must be between ", round(mincor[i,j], 3), " and ", round(maxcor[i,j], 3), "\n")
          errorCount = errorCount + 1
        }
      }
    }
  }
  cat("\n")
  if (errorCount > 0) {
    stop("Range violation occurred in the target correlation matrix!\n")
  }
  return(TRUE)
}
```


```{r}
# ValidCorrGpois(matrix(c(1, 0.9, 0.9, 1), byrow = TRUE, nrow = 2), c(0.5, 0.5), c(0.1, 0.105))
# ValidCorrGpois(matrix(c(1, 0.9, 0.9, 1), byrow = TRUE, nrow = 2), c(3, 2), c(-0.3, -0.2))
```


**QuantileGpois**

This function computes quantiles for generalized Poisson distribution. We guarantee that there will be at least five classes if lambda is negative by forcing $m \geq 4$.


```{r, echo = FALSE}
QuantileGpois = function(p, theta, lambda, details = FALSE){
  # Check if parameters are valid
  if(theta <= 0) stop("Theta has to be greater than 0!")
  if(lambda > 1) stop("Lambda has to be less than 1!")
  if(min(p) < 0 | max(p) > 1) stop("p must be between 0 and 1!")
  # since m >= 4, check lower bound for lambda when labda is negative
  if (lambda < 0 & lambda < (-theta)/4) stop (paste("For lambda<0, lambda must be >= -theta/4 = ", (-theta)/4, "!", sep=""))
  # Determine m
  m = numeric(1)
  if (lambda < 0) {
    mod = floor(-theta/lambda)
    if (-theta-mod * lambda == 0) m = mod
    else m = mod + 1
  }
  p.in = p
  upper = max(p.in)
  s = numeric(10000)
  q = numeric(length(p.in))
  w = exp(-lambda)
  #P(X = 0)
  p = exp(-theta)
  # s is for cumulative probability
  s[1] = p
  if (details) cat("x = 0, P(X = x) =",p, ",P(X <= x) =", s[1], "\n")
  i=1
  s[1] = as.numeric(p)
  if (details) cat("x = 0, P(X = x) =",p , ", P(X <= x) =", s[1], "\n")
  i = 1
  while (s[i] < upper) {
    #P(X = x)
    if (lambda < 0 & i > m) break
    else{
      p = theta * (theta + lambda * i)^(i - 1) * exp(-theta-lambda*i) / factorial(i)
      if(is.infinite(p)) {cat("error: p goes to infinite", "\n"); break  }
      if (i == 10000) {
        temp = numeric(10000)
        s = c(s, temp)
      }
        s[i + 1] = s[i] + p
        if (details) cat("x=", i, ", P(X = x) =", p, ", P(X <= x) =", s[i+1], "\n")
        i=i+1
     }
  }
  # For lambda < 0 to eliminate the truncation error ###
  if (lambda < 0) {
    Fm = s[i]
    s[1:i] = s[1:i] * Fm^(-1)
    if (details){
      cat("When lambda is negative, we need to account for truncation error\n")
      cat("The adjusted CDF are:", s[1:i])
      cat("\n")
    }
  }
  # quantile for x
  for (j in 1:length(p.in)) {
    i = 1
    while (p.in[j] > s[i]) {
      i = i + 1
    }
    q[j] = i - 1
  }
  return(q)
}
```


```{r}
QuantileGpois(0.98,1,-0.2,details = TRUE)
QuantileGpois(0.80,2,0.025,details = FALSE)
```


**CorrNNGpois**

This function applies the method proposed by Yahav, Shmueli 2011. Yahav and Shmueli found that the relationship between the desired correlation matrix and the actual correlation matrix of a generalized Poisson distribution can be approximated by an exponential function. Following their simple and empirically based approximation method we can correct our actual correlation matrix to the desired correlation matrix. Note that some desired correlations might be infeasible.


```{r}
CorrNNGpois = function(theta.vec, lambda.vec, r) {
  if(abs(r) > 1) stop("The desired correlation r has to be within (-1,1)")
  samples = 1e+05
  u = GenUniGpois(theta.vec[1], lambda.vec[1], samples, method = "Inversion", details = FALSE)$data
  v = GenUniGpois(theta.vec[2], lambda.vec[2], samples, method = "Inversion", details = FALSE)$data
  maxcor = cor(sort(u), sort(v))
  mincor = cor(sort(u), sort(v, decreasing=TRUE))
  a = -maxcor * mincor/(maxcor + mincor)
  b = log((maxcor + a)/a)
  c = -a
  corrected = log( (r + a)/a )/b
  if( abs(corrected) > 1 ) cat("The actual correlation, ",corrected,", is not feasible!", sep = "")
  else{ return(corrected) }
}
```


```{r}
CorrNNGpois(c(0.1,10), c(0.1, 0.2),0.5)
CorrNNGpois(c(0.1,10), c(-0.01, -0.02),0.5)
CorrNNGpois(c(4,2.3), c(-0.32,-0.3),0.7)
CorrNNGpois(c(14,10), c(-0.8, -0.3),0.99)
```


**CmatStarGpois**

This function computes the intermediate correlation values for Poisson-Poisson and Poisson-Normal pairs, and constructs an overall intermediate correlation matrix. It takes the target correlation matrix and returns the intermediate matrix of pairwise correlations.

The output of the `cmat.star` function is important because it is one of the input arguments for the main data generating function: `GenMVGpois`. The intermediate correlation matrix will lead to the target correlation matrix using inverse CDF transformation of the samples generated from a multivariate normal distribution.


```{r, echo = FALSE}
CmatStarGpois = function(corMat, theta.vec, lambda.vec){
  no.gpois = length(theta.vec)
  if (ValidCorrGpois(corMat, theta.vec, lambda.vec)) {
    corMat.star = diag(nrow(corMat))
    for (i in 1:nrow(corMat)) {
      for (j in 1:no.gpois) {
        if (i != j) {
          corMat.star[i,j] = CorrNNGpois(c(theta.vec[i], theta.vec[j]),
                                          c(lambda.vec[i], lambda.vec[j]),
                                          corMat[i, j])
        }
        cat(".")
      }
    }
  }
  cat("\n")
  if (!is.positive.definite(corMat.star)) {
    warning("Intermediate correlation matrix is not positive definite. Nearest positive definite matrix is used!")
    corMat.star = as.matrix(nearPD(corMat.star, corr = TRUE, keepDiag = TRUE)$mat)
  }
  corMat.star = (corMat.star + t(corMat.star))/2
  return(corMat.star)
}
```


```{r}
lambda.vec = c(-0.2, 0.2, -0.3)
theta.vec = c(1, 3, 4)
M = c(0.352, 0.265, 0.342)
N = diag(3)
N[lower.tri(N)] = M
TV = N + t(N)
diag(TV) = 1
cstar = CmatStarGpois(TV, theta.vec, lambda.vec)
cstar
```


**GenMVGpois**


`GenMVGPois` (the engine function) is the most important function in this package (**RNGforGPD**). It depends on all the other functions in this package and three external packages: **mvtnorm**, **corpcor**, and **VGAM**. The major difference between the univariate generalized Poisson variables generating function and the multivariate one is the consideration of pairwise correlations between variables. These correlations can be verified using `ValidCorrGpois` and corrected by `CorrNNGpois`.


```{r, echo = FALSE}
GenMVGpois = function(sample.size, no.gpois, cmat.star, theta.vec, lambda.vec, details = TRUE) {
  if (sample.size == 1 & details == TRUE) {
    stop("Parameter estimates are exactly the same as input of parameters in the case when n = 1!")
  }
  if (length(theta.vec) != no.gpois) {
    stop("Dimension of the theta vector does not match the number of generalized Poisson variables!\n")
  }
  if (length(lambda.vec) != no.gpois) {
    stop("Dimension of the lambda vector does not match the number of generalized Poisson variables!\n")
  }
  if (nrow(cmat.star) != no.gpois) {
    stop("Dimension of cmat.star and number of variables do not match!\n")
  }
  XX = rmvnorm(sample.size, rep(0, no.gpois), cmat.star)
  while (sum(pnorm(XX) > 0.99) != 0) {
    XX = rmvnorm(sample.size, rep(0, no.gpois), cmat.star)
  }
  YY = NULL
  for (i in 1:no.gpois) {
    UU = pnorm(XX[, i])
    XXgpois = QuantileGpois(UU, theta.vec[i],lambda.vec[i])
    YY = cbind(YY, XXgpois)
  }
  colnames(YY) = NULL
  if (sample.size != 1) {
    model = vglm(YY ~ 1, genpoisson(zero = 1))
    coef = matrix(Coef(model), 2, 3)
    emp.theta = round(coef[2, ], 6)
    emp.lambda = round(coef[1, ], 6)
    emp.corr = cor(YY)
    if (details == TRUE) {
      my.res = list(data = YY,
                  specified.theta = theta.vec,
                  empirical.theta = emp.theta,
                  specified.lambda = lambda.vec,
                  empirical.lambda = emp.lambda,
                  specified.corr = cmat.star,
                  empirical.corr = emp.corr)
      print(my.res[2:6])
      return(my.res)
    }
  }
  if (details == FALSE) {
    return(YY)
  }
}
```


```{r, warning = FALSE}
lambda.vec = c(-0.2, 0.2, -0.3)
theta.vec = c(1, 3, 4)
M = c(0.352, 0.265, 0.342)
N = diag(3)
N[lower.tri(N)] = M
TV = N + t(N)
diag(TV) = 1
cstar = CmatStarGpois(TV, theta.vec, lambda.vec)
sample.size = 2500; no.gpois = 3
data = matrix(NA, sample.size, no.gpois)
for (i in 1:25) {
     data[(((i - 1) * 100 + 1) : (i * 100)), ] = GenMVGpois(100, no.gpois, cstar, theta.vec, lambda.vec, details = FALSE)
}

apply(data, 2, mean) # empirical means
theta.vec / (1 - lambda.vec) # theoretical means
apply(data, 2, var) # empirical variances
theta.vec / (1 - lambda.vec)^3 # theoretical variances
cor(data) # empirical correlation matrix
TV # specified correlation matrix
```


# Citations

Amatya, A. and Demirtas, H. (2015). Simultaneous generation of multivariate mixed data with Poisson and normal marginals. \emph{Journal of Statistical Computation and Simulation}, \bold{85(15)}, 3129-3139.

Amatya, A. and Demirtas, H. (2017). PoisNor: An R package for generation of multivariate data with
Poisson and normal marginals. \emph{Communications in Statistics - Simulation and Computation}, \bold{46(3)}, 2241-2253.

Demirtas, H. (2017). On accurate and precise generation of generalized
Poisson variates. \emph{Communications in Statistics - Simulation and Computation},
\bold{46(1)}, 489-499.

Demirtas, H. and Hedeker, D. (2011). A practical way for computing approximate lower and upper correlation bounds. \emph{The American Statistician}, \bold{65(2)}, 104-109.

Yahav, I. and Shmueli, G. (2012). On generating multivariate Poisson data in management science applications. \emph{Applied Stochastic Models in Business and Industry}, \bold{28(1)}, 91-102.